---
title: "Bootstrap Bill"
author: "Myrthe Jager & Francis Blokzijl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
subtitle: Determine whether mutational profiles of two groups are significantly different
  using a bootstrap resampling method
---
# Distances between 96 mutational profiles

Samples that we have:

  * 15 Healthy liver 
  * 8 Alcoholic liver 

We want to test whether the 96 mutational profiles are significantly different between alcoholic and healthy liver (qualitatively). 

```{r include = FALSE}
# ---- GET STARTED ----

#1 Install & load required packages
library(MutationalPatterns)
library(ggplot2)
library(reshape2)

#2 Define directories
dir = ""
outdir = paste(dir,"Results/SNV/",sep = "")

# Read mutational profiles
mut_mat_all = read.table(paste(outdir,"mutational_profile.txt",sep=""), header = T)

# Load global variables from MutationalPatterns to get triplets for row.names
SUBSTITUTIONS = c('C>A','C>G','C>T','T>A','T>C','T>G')
SUBSTITUTIONS_96 = rep(SUBSTITUTIONS, each=16)

C_TRIPLETS = c(
    "ACA", "ACC", "ACG", "ACT",
    "CCA", "CCC", "CCG", "CCT",
    "GCA", "GCC", "GCG", "GCT",
    "TCA", "TCC", "TCG", "TCT")

T_TRIPLETS = c(
    "ATA", "ATC", "ATG", "ATT",
    "CTA", "CTC", "CTG", "CTT",
    "GTA", "GTC", "GTG", "GTT",
    "TTA", "TTC", "TTG", "TTT")

CONTEXTS_96 = c(rep(C_TRIPLETS, 3), rep(T_TRIPLETS, 3))

# combine substitutions and context in one 
TRIPLETS_96 = paste(substr(CONTEXTS_96,1,1), "[", SUBSTITUTIONS_96, "]", substr(CONTEXTS_96,3,3), sep = "")
remove(T_TRIPLETS,C_TRIPLETS,SUBSTITUTIONS,SUBSTITUTIONS_96,CONTEXTS_96)

# change rownames
rownames(mut_mat_all) = TRIPLETS_96

# Subset mut matrix for Healthy (WT) and Alcoholic (ALC) liver samples
mut_mat_WT_liver = mut_mat_all[,1:15]
mut_mat_ALC_liver = mut_mat_all[,16:23]


```

## 96 mutational profiles of the samples

```{r echo = F}

plot_96_profile(mut_mat_WT_liver, condensed = T, ymax = 0.1) + ggtitle("Healthy liver profiles")
plot_96_profile(mut_mat_ALC_liver, condensed = T, ymax = 0.1) + ggtitle("Alcoholic liver profiles")

```


```{r include = FALSE}
# FUNCTION A
# For each sample i (row) in mutation matrix m, generate n_rep replicas by sampling the same number of mutations with replacement
# Output is a mutation count matrix with n_rep * n_samples

generate_replicas = function(m, n_rep)
{
  replicas_list = list()
  # for each sample i
  for(i in 1:ncol(m))
  {
    profile = m[,i]
    replicas = matrix(nrow=96,ncol=n_rep)
    for(j in 1:n_rep)
    {
      # total number of mutations in sample
      n = sum(profile)
      # probability vectorof 96 trinucleotide changes
      p = profile/n
      # randomly take sample with replacement
      x = sample(TRIPLETS_96, size = n, prob = p, replace = T)
      # count the triplet appearances
      res = table(factor(x,lev=TRIPLETS_96))
      replicas[,j] = res
    }
    replicas_list[i] = list(replicas)
  }
  # combine replicas in a single matrix
  replicas_combined = do.call(cbind, replicas_list)
  return(replicas_combined)
}


# FUNCTION B
# Calculate centroid of matrix

centroid = function(m)
{
  m_centroid = rowSums(m)/ncol(m)
  return(m_centroid)
}

# FUNCTION C

profile_distance_permutations = 
  function(replicas_matrix_norm, n_samples, original_matrix_norm, n_perm, dist_measure)
{
  # Calculate centroid
  centroid_original = centroid(original_matrix_norm)
  
  dist_vector = c()
  for(y in 1:n_perm)
  {
    # Randomy select n_samples from the combined replicas matrix
    # Total number of replicates
    n_rep = ncol(replicas_matrix_norm)
    # Sample n_samples without replacement
    select = sample(1:n_rep, n_samples, replace = F)
    # Get profiles of these 3 samples
    replicas_selected_norm = replicas_matrix_norm[,select]
    # Calculate centroids
    centroid_replicas_selected = centroid(replicas_selected_norm)
    
    # Calculate distance of replicas centroid with centroid original matrix
    if(dist_measure == "euclidean")
    {
    # Calculate Euclidean distance
    res = dist(rbind(centroid_original, centroid_replicas_selected))
    }
    if(dist_measure == "cosine")
    {
     # Calculate cosine similarity with centroid of original samples
     res = 1 - cos_sim(centroid_original, centroid_replicas_selected)
    }
    dist_vector = c(dist_vector, res)
  }
  return(dist_vector)
}

# FUNCTION
# Plot bootstrap distribution
plot_density2 = function(distances_WT, distances_ALC, title, xlab, line, xlim)
{
  df = data.frame(Healthy = distances_WT, Alcoholic = distances_ALC)
  df = melt(df)
  # get values for P value = 0.05
  dist_001_Healthy = quantile(distances_WT, 0.99)
  dist_001_Alcoholic = quantile(distances_ALC, 0.99)
  ggplot(df, aes(x=value, group=variable, fill=variable)) +
    geom_density(alpha=.3) +
    ggtitle(title) +
    xlab(xlab) +
    expand_limits(x = 0) +
    xlim(c(0,xlim)) +
    theme_bw() +
    geom_vline(aes(xintercept=dist_001_Healthy), colour="red", linetype="dashed") +
    geom_vline(aes(xintercept=dist_001_Alcoholic), colour="lightblue", linetype="dashed") +
    geom_vline(aes(xintercept=line), colour="black")
}

```

\newpage

## Bootstrap distributions

* Generate bootstrap distribution for Healthy and Alcoholic liver samples
    + Generate in total 8x15x1000 = 120,000 replicas for each category

```{r include = FALSE}
# ---- GENERATE BOOTSTRAP DISTRIBUTION FOR WT LIVER SAMPLES ------

# generate replicas matrix for WT
mut_mat_WT_liver_rep = generate_replicas(mut_mat_WT_liver, 8000)
mut_mat_ALC_liver_rep = generate_replicas(mut_mat_ALC_liver, 15000)

```

# Differences between relative COSMIC signature contributions

* Generate bootstrap distribution for samples
    + Calculate centroid of the original WT samples
    + Determine the signature contribution
    + Randomly select number of profiles from replicas matrix (Healthy: 8, Alcoholic: 15)
    + Calculate centroid of the selected replicas
    + Determine the signature contribution 
    + Calculate distance between original and replicas normalized/relative signature contribution vectors
    + Do this 10000 times to get 10000 distances
    + Once with Cosine distance
    + Once with Euclidean distance

```{r include = FALSE}


# ----- FIT TO SIGNATURES -----

# FUNCTION
# Calculate distances between original signature contribution vector
# And contribution vector of 3 (combined) replicas

sig_contr_distance_permutations = function(replicas_matrix, n_samples, signatures, original_matrix, n_perm, dist_measure)
{
  # Get signature contributions for centroid of original matrix
  c1 = centroid(original_matrix)
  fit1 = fit_to_signatures(as.matrix(c1), signatures)
  original_contribution = as.vector(prop.table(fit1$contribution,2))
  
  # replicas
  dist_vector = c()
  for(y in 1:n_perm)
  {
    # Randomy select n_samples from the combined replicas matrix
    # Total number of replicates
    n_rep = ncol(replicas_matrix)
    # Sample n_samples without replacement
    select = sample(1:n_rep, n_samples, replace = F)
    # Get profiles of these samples
    replicas_selected = replicas_matrix[,select]
    # Calculate centroid or selected replicas
    c2 = centroid(replicas_selected)
    # Find signature contribution
    fit2 = fit_to_signatures(as.matrix(c2), signatures)
    # Normalize
    replicas_contribution = as.vector(prop.table(fit2$contribution,2))
 
    
    # Calculate distance of replicas centroid with centroid original matrix
    if(dist_measure == "euclidean normalized")
    {
      # Calculate Euclidean distance
      res = dist(rbind(original_contribution, replicas_contribution))
    }
    if(dist_measure == "cosine")
    {
      # Calculate cosine distance
      res = 1 - cos_sim(original_contribution, replicas_contribution)
    }
    dist_vector = c(dist_vector, res)
  }
  return(dist_vector)
} 

```

```{r include = F}

# ----- Read COSMIC mutational signatures ------
sp_url <- paste("http://cancer.sanger.ac.uk/cancergenome/assets/", "signatures_probabilities.txt", sep = "")
cancer_signatures = read.table(sp_url, sep = "\t", header = TRUE)
# Match the order of the mutation types to MutationalPatterns standard
new_order = match(TRIPLETS_96, cancer_signatures$Somatic.Mutation.Type)
# Reorder cancer signatures dataframe
cancer_signatures = cancer_signatures[as.vector(new_order),]
# Add trinucletiode changes names as row.names
row.names(cancer_signatures) = cancer_signatures$Somatic.Mutation.Type
# Keep only 96 contributions of the signatures in matrix
cancer_signatures = as.matrix(cancer_signatures[,4:33])
```

\newpage

```{r echo = F}

# Calculate signature contributions of centroids
sig_contr_30_WT_liver = fit_to_signatures(as.matrix(centroid(mut_mat_WT_liver)), cancer_signatures)$contribution
sig_contr_30_ALC_liver = fit_to_signatures(as.matrix(centroid(mut_mat_ALC_liver)), cancer_signatures)$contribution

# Normalize signature contributions
sig_contr_30_WT_liver_norm = prop.table(sig_contr_30_WT_liver,2)
sig_contr_30_ALC_liver_norm = prop.table(sig_contr_30_ALC_liver,2)

# Plot signature contribution of WT liver & ALC liver
df = cbind(sig_contr_30_WT_liver, sig_contr_30_ALC_liver)
colnames(df) = c("Healthy", "Alcoholic")
plot_contribution(df, coord_flip = TRUE) + ggtitle("Signature contributions liver")
plot_contribution_heatmap(df, cluster_samples = F) + ggtitle("Signature contributions liver")

# Calculate euclidean distance between signature contributions of WT liver & ALC liver
euc_dist_contr_WT_ALC_liver = dist(rbind(t(sig_contr_30_WT_liver_norm), t(sig_contr_30_ALC_liver_norm))) # 0.07878797

# Cosine
cos_dist_contr_WT_ALC_liver = 1 - cos_sim(as.vector(sig_contr_30_WT_liver_norm), as.vector(sig_contr_30_ALC_liver_norm)) # 0.01331735


```

```{r include = F}
# ------ Generate bootstrap distributions -------

# Euclidean
# liver
euc_dist_contr_WT_liver_perm = sig_contr_distance_permutations(mut_mat_WT_liver_rep,
                                       signatures = cancer_signatures,
                                       n_samples = 8,
                                       original_matrix = mut_mat_WT_liver,
                                       n_perm = 10000,
                                       dist_measure = "euclidean normalized")

euc_dist_contr_ALC_liver_perm = sig_contr_distance_permutations(mut_mat_ALC_liver_rep,
                                                         signatures = cancer_signatures,
                                                         n_samples = 15,
                                                         original_matrix = mut_mat_ALC_liver,
                                                         n_perm = 10000,
                                                         dist_measure = "euclidean normalized")
# Cosine
# liver
cos_dist_contr_WT_liver_perm = sig_contr_distance_permutations(replicas_matrix = mut_mat_WT_liver_rep,
                                                         signatures = cancer_signatures,
                                                         n_samples = 8,
                                                         original_matrix = mut_mat_WT_liver,
                                                         n_perm = 10000,
                                                         dist_measure = "cosine")

cos_dist_contr_ALC_liver_perm = sig_contr_distance_permutations(replicas_matrix = mut_mat_ALC_liver_rep,
                                                          signatures = cancer_signatures,
                                                          n_samples = 15,
                                                          original_matrix = mut_mat_ALC_liver,
                                                          n_perm = 10000,
                                                          dist_measure = "cosine")

```


```{r echo = F}

p1 = plot_density2(euc_dist_contr_WT_liver_perm, euc_dist_contr_ALC_liver_perm, 
              "Bootstrap distributions liver", 
              "Euclidean distance \n 30 signature contributions", 
              line = as.vector(euc_dist_contr_WT_ALC_liver),
              xlim = 0.4)

p1

p2 = plot_density2(cos_dist_contr_WT_liver_perm, cos_dist_contr_ALC_liver_perm, 
              "Bootstrap distributions liver", 
              "Cosine distance \n 30 signature contributions", 
              line = as.vector(cos_dist_contr_WT_ALC_liver),
              xlim = 0.4)

p2



pdf(paste(outdir,"bootstrap_30_liver.pdf",sep=""), useDingbats = F, height = 4)
p1
dev.off()


```
The dotted lines indicate the distances where P value = 0.01. The black line indicates the actual distance between the Healthy and Alcoholic samples. If this distance is larger than both distances with P value = 0.01, then we can say that the difference between the profiles is significantly different.

So the relative signature contribution vectors are not significantly different between Alcoholic and Healthy liver.